Particle Smoothing Algorithms for Variable Rate Models

### 2012-06-01 full_finaldraft_VRsmoothing_withSJGcomments.tex
moved models to section 2. lots of minor tidying.

Conditionally linear-Gaussian models for use with variable rate particle filters were introduced in \cite{Godsill2007a,Christensen2012} for financial and tracking problems. For example, changepoints might correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n^-)$, is inflated in an optimal fashion.

@@

Target tracking algorithms are commonly based upon fixed rate models (see, e.g. \cite{Li2003} for a thorough survey), in which the target kinematics (position, velocity, etc.) are estimated at a set of fixed times at which observations (e.g. radar measurements) are made. In \cite{Godsill2007a,Godsill2007,Whiteley2011}, variable rate models were introduced for tracking, in which the state trajectory is divided up by a set of changepoints between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division.

@@

with means and variances given by the following standard recursions (dependence on $\theta_{n}^-$ suppressed for clarity),
%
\begin{IEEEeqnarray}{rCl}
 m_n^- & = & A_n m_{n-1} \label{eq:kf_predict_start} \\
 P_n^- & = & A_n P_{n-1} A_n^T + Q_n \\
 \mu_n & = & C_n m_n^- \\
 S_n   & = & C_n P_n^- C_n^T + R_n \label{eq:kf_predict_stop} \\
 K_n   & = & P_n^- C_n^T S_n^{-1} \label{eq:kf_update_start}\\
 m_n   & = & m_n^- + K_n (y_n - \mu_n) \\
 P_n   & = & P_n^- - K_n S_n K_n^T \label{eq:kf_update_stop}    .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{-(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{-(i)})$.
    \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update_start})~to~(\ref{eq:kf_update_stop}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised variable rate particle filter}
\label{alg:RBVRPF}
\end{figure}

@@

The changepoint transition density may be expressed as,
%
\begin{IEEEeqnarray}{rCl}
 p(\theta_{n}^+|\theta_{n}^-) &=      & p(\tau_{K_n+1:K}, u_{K_n+1:K}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n) \nonumber \\
                                    &\propto& p(\tau_{K_n+1}, u_{K_n+1}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n)     .
\end{IEEEeqnarray}

The integral in~(\ref{eq:rb-vrps_back_cond_weight}) contains two terms involving the current state, $x_n$. The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict_start})--(\ref{eq:kf_update_stop}),
%
\begin{IEEEeqnarray}{rCl}
p(x_n|\theta_{n}^{-(i)}, y_{1:n}) & = & \mathcal{N}(x_n|m_n^{(i)}, P_n^{(i)})     .
\end{IEEEeqnarray}

@@

The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$. This may be regarded as an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}. Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF},
%
\begin{IEEEeqnarray}{rCl}
 p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
 p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \label{eq:backward_kf_predict_start} \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict_stop} \\
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \label{eq:backward_kf_update_start} \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
 \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}     .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
  \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}^-|y_{1:n})$ with weighted particles $\{\theta_{n}^{-(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{-(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$. Store all results.
  \FOR{$i=1 \dots N_S$}
  	\STATE Initialise particle using $\theta^{(i)} \sim \sum_j w_N^{(j)} \delta_{\theta^{(j)}}(\theta)$.
    \STATE Initialise sufficient statistics $\tilde{m}_N^{(i)}$ and $\tilde{P}_N^{(i)}$ (see text).
    \FOR{$n=N-1 \dots 1$}
      \STATE Discard $\theta_{n}^{-(i)}$.
      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using (\ref{eq:backward_kf_predict_start}) to (\ref{eq:backward_kf_predict_stop}).
      \FOR{$j=1 \dots N_P$}
	      \STATE Calculate weight $\tilde{w}_n^{(j)}$ using (\ref{eq:rb-vrps_back_cond_weight2}).
      \ENDFOR
      \STATE Sample $\theta_{n}^{-(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{-(j)}}(\theta_{n}^-)$.
      \STATE Join $\theta^{(i)} \gets \theta_{n}^{-(i)} \cup \theta_{n}^{+(i)}$.
      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using (\ref{eq:backward_kf_update_start}) to (\ref{eq:backward_kf_update_stop}).
    \ENDFOR
  \ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised Variable Rate Particle Smoother}
\label{alg:RBVRPS}
\end{figure}

@@

In order to achieve conditional independence between past and future observations, consider the augmented changepoint sequence, $\tilde{\theta} = \{\tau_{0:K}, u_{0:K}, x_{0:K}\}$ ($\tilde{\theta}_n^-$ and $\tilde{\theta}_n^+$ defined in the same manner as before). The changepoint state values, $x_{0:K}$ may be calculated deterministically from $\theta$ for each particle. The smoother is then formulated by factorising the augmented sequence posterior distribution as before,
%
\begin{IEEEeqnarray}{rCl}
 p(\tilde{\theta}|y_{1:N}) = p(\tilde{\theta}_{n}^{+}|y_{1:N}) p(\tilde{\theta}_{n}^-|\tilde{\theta}_{n}^{+}, y_{1:N})     .
\end{IEEEeqnarray}

@@

For this method to work, a new condition must be imposed on the state dynamics. For any pair of adjacent changepoints, $(\tau_k,x_k)$, and $(\tau_{k+1},x_{k+1})$ it must be possible to calculate a parameter value, $u_k$, which results in the transition from the former to the latter. Generally, this requires there to be as many degrees of freedom in the selection of $u_k$ as there are state dimensions of $x_k$.

@@

Each Markov chain may be initialised with the output from the previous stage. This value is itself a particle from the target distribution (albeit one from an approximation with poor path-space diversity), meaning that no burn-in period is required. This is the same reasoning which justifies the absence of a burn-in when using a resample-move scheme \cite{Gilks2001}.

@@

The Rao-Blackwellised VRPS algorithm was tested on the financial time series model of \cite{Christensen2012}, in which prices of an asset are treated as noisy observations of a latent state, which evolves according to a drift-diffusion with occasional jumps.

The latent state is a vector with two elements, the underlying value of the asset, and the trend followed by this value,
%
\begin{equation}
 \mathbf{x}(t) = [ x(t), \dot{x}(t)]^T     .
\end{equation}

This evolves continuously according to a jump-diffusion model,
%
\begin{IEEEeqnarray}{c}
 d\mathbf{x}(t) = \begin{bmatrix}0 & 1 \\ 0 & -\lambda \end{bmatrix} \mathbf{x}(t) dt + \begin{bmatrix}0 \\ \sigma \end{bmatrix} d\beta(t) + d\mathbf{J}(t)
\end{IEEEeqnarray}

where $\lambda$ introduces a mean reversion effect on the trend and $\mathbf{\beta}(t)$ is standard Brownian motion (with unit diffusion constant). The jump term, $d\mathbf{J}(t)$, is zero everywhere except when jumps occur,
%
\begin{IEEEeqnarray}{rCl}
 d\mathbf{J}(t) & = & \begin{cases} \mathbf{J}_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
 \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})     .
\end{IEEEeqnarray}

This model may be discretised at the observation times using standard methods, detailed in appendix~\ref{app:lg_model_discretisation}, leading to,% (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
%
%Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
%
\begin{IEEEeqnarray}{rCl}
 \mathbf{x}_n &=& A \mathbf{x}_{n-1} + \mathbf{w}_n \\
 y_n          &=& C \mathbf{x}_{n} + v_n
\end{IEEEeqnarray}

where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively. $Q_n$ is a function of the sequence of jump times and their types, giving rise to the desired conditionally linear-Gaussian structure.% The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
%
%\begin{IEEEeqnarray}{rCl}
% A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
% C     &=& \begin{bmatrix}1 & 0\end{bmatrix}
%\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
% Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
%                        Q_{D}             & \text{otherwise} \end{cases} \\
% Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
% q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
% q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
% q_3   &=& 1-e^{(-2 \lambda \Delta t)} \\
% R     &=& [\sigma_y^2]
%\end{IEEEeqnarray}

The times between changepoints are assumed to be exponentially distributed with mean $1/\alpha$, and equal probability of value and trend jumps.
%
%\begin{IEEEeqnarray}{rCl}
% \IEEEeqnarraymulticol{3}{l}{p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) = P(u_k) p(\tau_k|\tau_{k-1})} \\
% p(\tau_k|\tau_{k-1})               &=& \begin{cases}\frac{1}{\alpha} \exp ( \alpha (\tau_k-\tau_{k-1}) ) & \tau_k>\tau_{k-1} \\
%                                                     0 & \tau_k < \tau_{k-1} \end{cases} \\
% P(u_k)                             &=& \begin{cases}0.5 & u_k = 1 \\ 0.5 & u_k = 2\end{cases}
%\end{IEEEeqnarray}

The algorithms were tested on artificial data simulated from this model. Observations were made on a regular time grid with a spacing of $0.0017s$. Other parameters used were: $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.

% The decay constant for the distribution of inter-changepoint times is 20, and the trend mean-reversion coefficient is 5. The diffusion constant is 0.05 and the observation noise standard deviation 0.001. The jump standard deviations are 0.005 and 0.05 respectively for value and trend. (Terms defined in appendix~\ref{app:lg_model_discretisation})

%The following parameters were used: $\Delta t = 0.0017$, $N = 1000$, $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.

@@

%For a first, qualitative comparison of the changepoint sequence distributions, it is possible to generate kernel density estimates for the changepoint times. These are constructed by simply adding a (unit amplitude) Gaussian kernel for each changepoint present in each particle and scaling by the number of particles. The resulting function provides an approximate measure of the likelihood of finding a jump at a given time. These are shown in figure~\ref{fig:example_kdest} for the example data set.

%\begin{figure}[!t]
%\centering
%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_kdest.pdf}} \\
%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_kdest.pdf}}
%\caption{Filter (a) and smoother (b) kernel density estimates for value (top) and trend (bottom) jump times. Correct times overlayed as dotted vertical lines.}
%\label{fig:example_kdest}
%\end{figure}

%It is immediately apparent that the approximation produced by the filter lacks diversity. The jump time kernel density estimate comprises peaks with magnitude $1$, with regions of $0$ between. This is a result of every particle containing an almost identical list of changepoints, as a result of resampling. In contrast, the smoother is clearly providing a less degenerate particle representation, and generally contains larger kernel density peaks at the times of larger jumps. Furthermore, the kernel density for the trend jump times contains broader peaks, capturing the fact that because the trend is not observed directly, the jumps are harder to localise precisely.

%A further comparison of the algorithms is possible via the state estimates they generate. Here we can compare three options: the filtering results, using the VRPF and a Kalman filter for the state estimates; the filter-smoother results, using the final VRPF approximation for the changepoint sequence and a Rauch-Tung-Striebel (RTS) smoother for the state estimates; and the full smoothing results, using the VRPS followed by an RTS smoother. For the example data set, the trend estimates are shown in figure~\ref{fig:example_state} (The estimates of value are less informative, as this quantity is observed). Again, the improved particle diversity of the smoother is apparent.

%\begin{figure}[!t]
%\centering
%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_state.pdf}} \\
%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filtersmoother_state.pdf}} \\
%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_state.pdf}}
%\caption{Trend estimates using RBVRPF + Kalman filter (a) RBVRPF + RTS smoother (b) and RBVRPS + RTS smoother. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
%\label{fig:example_state}
%\end{figure}

>>>>

\subsubsection*{A Jump-Diffusion model for Financial Assets} \label{sec:financial_model}

Conditionally linear-Gaussian models for use with variable rate particle filters were introduced in \cite{Godsill2007a,Christensen2012} for financial applications. A version of these models is used for testing the new smoothing algorithm. Prices of an asset are treated as noisy observations of a latent state, which evolves according to a drift-diffusion with occasional jumps.

The state is a vector with two elements, the underlying value of the asset, and the trend followed by this value,
%
\begin{equation}
 x(t) = [ z(t), \dot{z}(t)]^T     .
\end{equation}

This evolves continuously according to a jump-diffusion model,
%
\begin{IEEEeqnarray}{c}
 dx(t) = \begin{bmatrix}0 & 1 \\ 0 & -\lambda \end{bmatrix} x(t) dt + \begin{bmatrix}0 \\ \sigma \end{bmatrix} d\beta(t) + dJ(t)
\end{IEEEeqnarray}

where $\lambda$ introduces a mean reversion effect on the trend and $\beta(t)$ is standard Brownian motion (with unit diffusion constant). The jump term, $dJ(t)$, is zero everywhere except when jumps occur,
%
\begin{IEEEeqnarray}{rCl}
 dJ(t) & = & \begin{cases} J_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
 J_k  & \sim & \mathcal{N}(J_k| 0, Q_{J,u_k})     .
\end{IEEEeqnarray}

This model may be discretised at the observation times using standard methods, detailed in appendix~\ref{app:lg_model_discretisation}, leading to,% (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
%
%Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
%
\begin{IEEEeqnarray}{rCl}
 x_n & = & A x_{n-1} + w_n \\
 y_n & = & C x_{n} + v_n
\end{IEEEeqnarray}

where the $w_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively. $Q_n$ is a function of the sequence of jump times and their types, giving rise to the desired conditionally linear-Gaussian structure.

@@

\subsubsection*{An Intrinsic Coordinate Model for Manoeuvring Vehicles} \label{sec:tracking_model}

Target tracking algorithms are commonly based upon fixed rate models (see, e.g. \cite{Li2003} for a thorough survey), in which the target kinematics (position, velocity, etc.) are estimated at a set of fixed times at which observations (e.g. radar measurements) are made. In \cite{Godsill2007a,Godsill2007,Whiteley2011,Bunch2012a}, variable rate models were introduced for tracking. The state trajectory is divided up by a set of changepoints, between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division. In this paper we use the 2-dimensional intrinsic coordinate model of \cite{Bunch2012a} to test the new smoothing algorithm.

The target is modelled as a particle subject to perpendicular forces tangential and normal to the direction of motion. These forces are held constant between changepoint times, $\tau_{1:K}$, which represent the beginnings and ends of manoeuvres. The state vector consists of Cartesian position, $x_1(t)$ and $x_2(t)$, plus heading angle, $\psi(t)$, and speed, $\dot{s}(t)$,
%
\begin{equation}
x(t) = [x_1(t), x_2(t), \psi(t), \dot{s}(t)]^T     .
\end{equation}

In addition to the tangential and normal accelerations, $a_{T,k}$ and $a_{N,k}$, we introduce two linear velocity terms, $d_{X,k}$ and $d_{Y,k}$. As will become apparent later, these are necessary for the operation of an efficient smoothing algorithm. They provide us with enough degrees of freedom to match the state trajectory between an arbitrary past changepoint ($\tau_{K(t)}$, $\mathbf{x}_{K(t)}$) and an arbitrary future changepoint ($\tau_{K(t)+1}$, $\mathbf{x}_{K(t)+1}$). These linear velocity parameters may be considered to be merely relaxation terms, or they may represent real physical effects, such as wind (for aircraft) or currents (for boats). Together, these four variables make up the vector of motion parameters,
%
\begin{equation}
u_k = [a_{T,k}, a_{N,k}, d_{X,k}, d_{Y,k}]^T     .
\end{equation}

The target dynamics are described by four differential equations,
%
\begin{IEEEeqnarray}{rCl}
\ddot{s}(t) & = & a_{T,K(t)} \\
\dot{s}(t) \dot{\psi}(t) & = & a_{N,K(t)} \\
\dot{x}_{1}(t) & = & \dot{s}(t) \cos(\psi(t)) + d_{X,K(t)} \\
\dot{x}_{2}(t) & = & \dot{s}(t) \sin(\psi(t)) + d_{Y,K(t)}     .
\end{IEEEeqnarray}

Solving these yields the following state equations (where $\Delta\ t = t - \tau_{K(t)}$),
%
\begin{IEEEeqnarray}{rCl}
\dot{s}(t) & = & \dot{s}_{K(t)} + a_{T,k} \Delta t \label{eq:2D_ICmodel_1} \\
\psi(t) & = & \psi_{K(t)} + \frac{a_{N,k}}{a_{T,k}} \log \left( \frac{\dot{s}(t)}{\dot{s}_{K(t)}} \right) \label{eq:2D_ICmodel_2}
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
x_1(t) & = & x_{1,K(t)} + d_{X,K(t)} \Delta t \label{eq:2D_ICmodel_3} \\
     \IEEEeqnarraymulticol{3}{l}{ \quad + \: \frac{ \dot{s}(t)^2 }{ 4 a_{T,k}^2 + a_{N,k}^2 } \left[  a_{N,k} \sin(\psi(t)) + 2 a_{T,k} \cos(\psi(t))  \right]} \nonumber \\
     \IEEEeqnarraymulticol{3}{l}{ \quad - \: \frac{\dot{s}_{K(t)}^2}{4 a_{T,k}^2 + a_{N,k}^2} \left[  a_{N,k} \sin(\psi_{K(t)}) + 2 a_{T,k} \cos(\psi_{K(t)})  \right]} \nonumber
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
x_2(t) & = & x_{2,K(t)} + d_{Y,K(t)} \Delta t \label{eq:2D_ICmodel_4} \\
     \IEEEeqnarraymulticol{3}{l}{ \quad + \: \frac{ \dot{s}(t)^2 }{ 4 a_{T,k}^2 + a_{N,k}^2 } \left[ -a_{N,k} \cos(\psi(t)) + 2 a_{T,k} \sin(\psi(t))  \right]} \nonumber \\
     \IEEEeqnarraymulticol{3}{l}{ \quad - \: \frac{\dot{s}_{K(t)}^2}{4 a_{T,k}^2 + a_{N,k}^2} \left[  -a_{N,k} \cos(\psi_{K(t)}) + 2 a_{T,k} \sin(\psi_{K(t)})  \right]} \nonumber      .
\end{IEEEeqnarray}

When formulating the smoothing algorithm, it will be necessary that $u_k$ can be calculated for any pair of adjacent changepoints, $(\tau_k,x_k)$, and $(\tau_{k+1},x_{k+1})$. This condition is met by the preceding model --- the equations are solved in the order specified to find $a_{T,k}$ from (\ref{eq:2D_ICmodel_1}), $a_{N,k}$ from (\ref{eq:2D_ICmodel_2}), and finally $d_{X,K(t)}$ and $d_{Y,K(t)}$ from (\ref{eq:2D_ICmodel_3}) and (\ref{eq:2D_ICmodel_4}).

@@

with means and variances given by the following standard recursions (dependence on $\theta_{n}^-$ suppressed for clarity),
%
\begin{IEEEeqnarray}{rCl}
 m_n^- & = & A_n m_{n-1} \nonumber \\
 P_n^- & = & A_n P_{n-1} A_n^T + Q_n \nonumber \\
 \mu_n & = & C_n m_n^- \nonumber \\
 S_n   & = & C_n P_n^- C_n^T + R_n \label{eq:kf_predict} \\
 \nonumber \\
 G_n   & = & P_n^- C_n^T S_n^{-1} \nonumber \\
 m_n   & = & m_n^- + G_n (y_n - \mu_n) \nonumber \\
 P_n   & = & P_n^- - G_n S_n G_n^T \label{eq:kf_update}    .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{-(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
\FOR{$n=1 \dots N$}
  \FOR{$i=1 \dots N_F$}
  	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{-(i)})$.
    \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict}).
    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update}).
  \ENDFOR
  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
\ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised variable rate particle filter}
\label{alg:RBVRPF}
\end{figure}

@@

When calculating the changepoint transition density of the future sequence, only the terms depending on the past sequence need to be considered, because the future sequence is fixed,
%
\begin{IEEEeqnarray}{rCl}
 p(\theta_{n}^+|\theta_{n}^-) & =     & p(\tau_{K_n+1:K}, u_{K_n+1:K}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n) \nonumber \\
                              &\propto& p(\tau_{K_n+1}, u_{K_n+1}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n)     .
\end{IEEEeqnarray}

The integral in~(\ref{eq:rb-vrps_back_cond_weight}) contains two terms involving the current state, $x_n$. The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict}) and (\ref{eq:kf_update}),
%
\begin{IEEEeqnarray}{rCl}
p(x_n|\theta_{n}^{-(i)}, y_{1:n}) & = & \mathcal{N}(x_n|m_n^{(i)}, P_n^{(i)})     .
\end{IEEEeqnarray}

@@

The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$. This may be regarded as an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}. Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF},
%
\begin{IEEEeqnarray}{rCl}
 p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
 p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict} \\
 \nonumber \\
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \nonumber \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \nonumber \\
 \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \label{eq:backward_kf_update}     .
\end{IEEEeqnarray}

@@

\begin{figure}
\fbox{\parbox{\columnwidth}{
\begin{algorithmic}[1]
\linespread{1.5} \selectfont
  \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}^-|y_{1:n})$ with weighted particles $\{\theta_{n}^{-(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{-(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$. Store all results.
  \FOR{$i=1 \dots N_S$}
  	\STATE Initialise particle using $\theta^{(i)} \sim \sum_j w_N^{(j)} \delta_{\theta^{(j)}}(\theta)$.
    \STATE Initialise sufficient statistics $\tilde{m}_N^{(i)}$ and $\tilde{P}_N^{(i)}$ (see text).
    \FOR{$n=N-1 \dots 1$}
      \STATE Discard $\theta_{n}^{-(i)}$.
      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using (\ref{eq:backward_kf_predict}).
      \FOR{$j=1 \dots N_F$}
	      \STATE Calculate weight $\tilde{w}_n^{(j)}$ using (\ref{eq:rb-vrps_back_cond_weight2}).
      \ENDFOR
      \STATE Sample $\theta_{n}^{-(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{-(j)}}(\theta_{n}^-)$.
      \STATE Join $\theta^{(i)} \gets \theta_{n}^{-(i)} \cup \theta_{n}^{+(i)}$.
      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using (\ref{eq:backward_kf_update}).
    \ENDFOR
  \ENDFOR
\end{algorithmic}
}}
\caption{Rao-Blackwellised Variable Rate Particle Smoother}
\label{alg:RBVRPS}
\end{figure}

@@

In order to achieve conditional independence between past and future observations, consider the augmented changepoint sequence, $\tilde{\theta} = \{\tau_{0:K}, u_{0:K}, x_{0:K}\}$ ($\tilde{\theta}_n^-$ and $\tilde{\theta}_n^+$ defined in the same manner as before). The changepoint state values, $x_{0:K}$ may be calculated deterministically from $\theta$ for each particle. The smoother is then formulated by factorising the augmented sequence posterior distribution,
%
\begin{IEEEeqnarray}{rCl}
 p(\tilde{\theta}|y_{1:N}) = p(\tilde{\theta}_{n}^{+}|y_{1:N}) p(\tilde{\theta}_{n}^-|\tilde{\theta}_{n}^{+}, y_{1:N})     .
\end{IEEEeqnarray}

@@

For this method to work, a new condition must be imposed on the state dynamics. For any pair of adjacent changepoints, $(\tau_k,x_k)$, and $(\tau_{k+1},x_{k+1})$ it must be possible to calculate a parameter value, $u_k$, which results in the transition from the former to the latter. Generally, this requires there to be as many degrees of freedom in the selection of $u_k$ as there are state dimensions of $x_k$. This condition is satisfied by the tracking model of section~\ref{sec:tracking_model} only through the introduction of the linear drift parameters.

@@

Each Markov chain may be initialised with the output from the previous stage. This value is itself a particle from the target distribution (albeit one from an approximation with poor path-space diversity), meaning that no burn-in period is required. However, longer chains are likely to result in less dependence between samples and a reduction in Monte Carlo error. This is the same reasoning which justifies the absence of a burn-in when using a resample-move scheme \cite{Gilks2001}.

@@

The Rao-Blackwellised variable rate algorithms were tested on the jump-diffusion model outlined in section~\ref{sec:financial_model} using artificial simulated data. The times between changepoints are assumed to be exponentially distributed with mean $1/\alpha$, and equal probability of value and trend jumps. Observations were made on a regular time grid with a spacing of $0.0017s$. Other parameters used were: $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.

@@
